```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(mice)
library(dplyr)
library(naniar)
library(survival)
library(lme4)
```


```{r}
heart_data <- read.csv(file.choose())
```


# EDA

## Baselline table

```{r}
# pick baseline vars (edit if you want)
key_vars <- heart_data %>%
  select(sex, age, time, fuyrs, status, grad, log.lvmi, ef, lvh, prenyha, redo, size, con.cabg, creat, dm, acei, lv, emergenc, hc, sten.reg.mix, hs)

baseline_summary <- key_vars %>%
  group_by(hs) %>%
  summarise(across(everything(),
                   list(mean = ~mean(. , na.rm = TRUE),
                        sd = ~sd(. , na.rm = TRUE))))

baseline_summary
```

```{r}
# Adding patient data so pre surgery measurements arent duplicated for plots
patient_data <- heart_data[!duplicated(heart_data$num), ]
patient_data
```



## Boxplots of selected baselines covariates by survival status 

```{r}
# remove this one for bar plot
ggplot(heart_data, aes(x = factor(status), y = sex)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Sex")

ggplot(heart_data, aes(x = factor(status), y = age)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Age")

ggplot(heart_data, aes(x = factor(status), y = grad)) +
  geom_boxplot() + 
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Valve gradient at follow-up")

ggplot(subset(heart_data, grad < 30), 
       aes(x = factor(status), y = grad)) +
  geom_boxplot() + 
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Valve gradient at follow-up")

ggplot(heart_data, aes(x = factor(status), y = lvmi)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Left ventricular mass index")

ggplot(subset(heart_data, lvmi < 300), aes(x = factor(status), y = lvmi)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Left ventricular mass index")

ggplot(heart_data, aes(x = factor(status), y = ef)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Ejection fraction at follow-up visit")

ggplot(patient_data, aes(x = factor(status), y = bsa)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative body surface area")

ggplot(patient_data, aes(x = factor(status), y = size)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "size of the valve (millimeters)")

ggplot(patient_data, aes(x = factor(status), y =  creat)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative serum creatinine")

ggplot(subset(patient_data, creat < 200), aes(x = factor(status), y =  creat)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative serum creatinine")

ggplot(heart_data, aes(x = factor(status), y =  creat)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative serum creatinine")
```
- age 
- lmvi
- bsa

## Bar plots of Survival rates vs categorical variables

NOTE: thinking of using the wide form data for these so that patients don't stack
```{r}
# hs + sex + age + bsa + lvh + prenyha + redo + dm + lv + acei + hs + sten.reg.mix

# Survivors vs dead
ggplot(patient_data, aes(x = as.logical(status), fill = as.logical(status))) +
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5, size=5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_x_discrete(labels = c("Survived", "Died")) +
  scale_fill_manual(values = c("orange", "darkgrey"), guide = "none") + 
  labs(title = "Status",
       x = "Status",
       y = "Count")

# sex vs survival status
ggplot(patient_data, aes(x = factor(sex, labels = c("Male", "Female")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Valve Type", title = "Survival by Sex") +
  scale_fill_manual(values = c("orange", "darkgrey"))

# graft type vs survival status
ggplot(patient_data, aes(x = factor(hs, labels = c("Stentless", "Homograft")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Valve Type", title = "Survival by Valve Type") +
  scale_fill_manual(values = c("orange", "darkgrey"))

# Sten.Reg.Mix vs survival status
ggplot(patient_data, aes(x = factor(sten.reg.mix, labels = c("Stenosis", "Regurgitation", "Mixed")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Haemodynamics", title = "Survival by Valve Pathology") +
  scale_fill_manual(values = c("orange", "darkgrey"))

# LV vs survival status
ggplot(patient_data, aes(x = factor(lv, labels = c("Good", "Moderate", "Poor")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Pre-op LV Function", title = "Survival by Heart Function") +
  scale_fill_manual(values = c("orange", "grey"))

# Diabetes vs survival status
ggplot(patient_data, aes(x = factor(dm, labels = c("No", "Yes")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Diabetes", title = "Survival by Diabetes") +
  scale_fill_manual(values = c("orange", "grey"))

# Severity vs survival status
ggplot(patient_data, aes(x = factor(prenyha, labels = c("Class I/II", "Class III/IV")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Diabetes", title = "Survival by severity") +
  scale_fill_manual(values = c("orange", "grey"))

# redone surgery vs survival status
ggplot(patient_data, aes(x = factor(redo, labels = c("No", "Yes")), fill = as.logical(status))) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", x = "Valve Type", title = "Survival by Redone Surgery") +
  scale_fill_manual(values = c("orange", "darkgrey"))

```

## Kaplan-Meier curves by valve type

```{r}
# create a survival object
# fuyrs = follow-up time
surv_obj <- Surv(heart_data$fuyrs, heart_data$status)

# this estimates a separate survival curve for each valve group
fit <- survfit(surv_obj ~ hs, data = heart_data)

plot(fit,
     col = c("blue", "red"),
     xlab = "Time since surgery (years)",
     ylab = "Survival probability")

legend("bottomleft",
       legend = c("Porcine", "Homograft"),
       col = c("blue", "red"),
       lty = 1)
```
Potential differences in survival between valve types, with the homograft group showing lower survival probabilities over follow-up. This motivates inclusion of valve type in the survival submodel

```{r}
# sort so first row = earliest time
heart_sorted <- heart_data %>%
  arrange(num, time)

# build subject-level status 
surv_status <- heart_sorted %>%
  distinct(num, .keep_all = TRUE) %>% # earliest row per subject
  transmute(num,
            status_f = factor(as.integer(status),
                              levels = c(0, 1),
                              labels = c("Lost at follow-up", "Died"))
            )

# merge back to full longitudinal data
plot_data <- heart_data %>%
  left_join(surv_status, by = "num") %>%
  filter(!is.na(status_f),
         !is.na(time),
         !is.na(log.lvmi))
```

# compare logl(lvmi) rate of change by survival status

```{r,warning=FALSE,message=FALSE}
ggplot(heart_data,
       aes(time, log.lvmi,
           color = factor(status))) +
  geom_smooth(se=FALSE)
```

The curves are not dramatically different in slope. There is some early
divergence, but overall the shapes are similar. While there is heterogeneity in trajectories, there is no clear systematic difference in rate of change between
groups.

# smoothed mean log(lvmi) over time by survival status

```{r,warning=FALSE,message=FALSE}
mean_data <- plot_data %>%
  group_by(status_f, time) %>%
  summarise(mean_log_lvmi = mean(log.lvmi, na.rm = TRUE),
            n = n(),
            .groups = "drop")

ggplot(mean_data,
       aes(x = time,
           y = mean_log_lvmi,
           group = status_f,
           linetype = status_f)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = TRUE,
              method = "loess",
              span = 0.75) +
  labs(title = "Smoothed Mean log(LVMI) Over Time by Survival Status",
       x = "Time",
       y = "Mean log(LVMI)",
       linetype = "Survival Status")

ggplot(plot_data,
       aes(x = time,
           y = log.lvmi,
           group = status_f,
           linetype = status_f)) +
  geom_smooth(se = TRUE,
              method = "loess",
              span = 0.75) +
  labs(title = "LOESS-Smoothed log(LVMI) Trajectories by Survival Status",
       x = "Time",
       y = "log(LVMI)",
       linetype = "Survival Status")
```

1st: 

There is no dramatic divergence in trajectory shape. Patients who died tend to
maintain slightly higher LVMI levels over time. Mortality risk might be
associated with higher LVMI level.

2nd:

Both groups show early postoperative decline. The Lost group trends slightly
upward over time. The Died group remains somewhat elevated but does not show
clearly steeper decline. At later time points, the Died curve becomes unstable
(very wide confidence band). Differences are mostly in absolute level, not in
trajectory slope. No clear systematic difference in rate of change.

# distribution of log(lvmi) by survival status

```{r}
ggplot(plot_data,
       aes(x = status_f,
           y = log.lvmi)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15,
              alpha = 0.3) +
  labs(title = "Distribution of log(LVMI) by Survival Status",
       x= "Survival Status",
       y = "log(LVMI)")
```

The Died group has a visibly higher median. The distribution is shifted upward.
There is overlap, but the central tendency differs. Patients who died tended to

have higher LVMI values overall.
