---
title: "STAT 293 - Final Project"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(ggplot2)
library(naniar)
library(survival)
library(lme4)
library(dplyr)
```


```{r}
heart_data <- read.csv(file.choose())
```

# ANALYSIS DRAFT

```{r}
lmm_fit <- lmer(
  log.lvmi ~ time * hs + sex + age + bsa + lvh + prenyha + redo + dm + lv + (time | num),
  data = heart_data)

summary(lmm_fit)
```

WHAT I GOT FOR LMM:
- interaction: 0.005 No evidence that rate of LVMI change differs by valve type
- time: -0.012 LVMI slighlty decreases over time after surgery 
- hs: 0.068 Patients with stentless valves have slightly higher LVMI, but not statistically convincing.
- sex: -0.218 Females have lower LVMI on average
- age: 0.001 not significant
- bsa: -0.274 Larger BSA is associated with lower indexed LVMI
- lvh: -0.114 patients with preoperative LVH have slightly lower postoperative LVMI
- prenyha: -0.005 not significant
- redo: -0.007 Patients with prior surgery have slightly higher LVMI, but effect is small
- dm: -0.191 slightly lower LVMI among diabetes 
- lv: 0.038 higher lvef means higher lmvi

```{r}
survival_fit <- coxph(
  Surv(fuyrs, status) ~ hs +
    sex + age + bsa + lvh + prenyha +
    redo + dm + lv,
  data = heart_data
)

summary(survival_fit)
```

WHAT I GOT FOR SURVIVAL ANALYSIS:
- hr: 2.02 patients with stentless valves have about 2× hazard of death
- age: 1.18 each additional year of age means 18% higher hazard 
- lvh: 0.14 
- redo 0.494 prior surgery associated with lower observed hazard
- lv: 2.49 worse LV function means higher mortality risk


# ANALYSIS BASED OF ARTICLE WITH JM

```{r}
#install.packages("nlme")
library(nlme)
#install.packages("JM")
library(JM)

jm_data <- heart_data[!duplicated(heart_data$num), ]

lme_fit <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + prenyha + redo + dm + lv,
  random = ~ time | num,
  data = heart_data)

cox_fit <- coxph(
  Surv(fuyrs, status) ~ hs + sex + age + bsa + lvh + prenyha + redo + dm + lv +
    cluster(num),
  data = jm_data,
  x = TRUE,
  model = TRUE
)


joint_fit <- jointModel(
  lme_fit,
  cox_fit,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)

summary(joint_fit)
```

# slope parameterization

```{r}
#reload the full dataset
heart_data <- read.csv(file.choose())

# sort long data 
heart_sorted <- heart_data %>%
  arrange(num, time)

# build subject-level survival data from earliest row per subject
jm_data <- heart_sorted %>%
  distinct(num, .keep_all = TRUE) %>% # earliest time after sorting
  mutate(status = as.integer(status)) %>%
  filter(!is.na(fuyrs),
         fuyrs > 0,
         !is.na(status))

# longitudinal data for JM: drop subject-level survival columns to avoid .x/.y 
heart_long <- heart_sorted %>%
  dplyr::select(-any_of(c("fuyrs", "status"))) %>% # remove if exist in long file
  distinct(num, time, .keep_all = TRUE) # remove duplicate times

# enforce measurement times within follow-up 
heart_jm <- heart_long %>%
  inner_join(jm_data %>% dplyr::select(num, fuyrs),
             by = "num") %>%
  filter(time <= fuyrs) %>% # strict to avoid stop = start
  dplyr::select(-fuyrs)

# restrict to obs that complete data for both models
lmm_vars <- c("num", "time", "log.lvmi", "hs", "sex", "age", "bsa", "lvh", 
              "prenyha", "redo", "dm", "lv")
ids_lmm <- heart_jm %>%
  dplyr::select(all_of(lmm_vars)) %>%
  tidyr::drop_na() %>%
  distinct(num) %>%
  pull(num)

cox_vars <- c("num", "fuyrs", "status",  "hs", "sex", "age",   "bsa", "lvh",
              "prenyha", "redo", "dm", "lv")
ids_cox <- jm_data %>%
  dplyr::select(all_of(cox_vars)) %>%
  tidyr::drop_na() %>%
  distinct(num) %>%
  pull(num)

keep_ids <- intersect(ids_lmm, ids_cox)
heart_jm2 <- heart_jm %>% filter(num %in% keep_ids)
jm_data2 <- jm_data %>% filter(num %in% keep_ids)

# refit models on cleaned datasets
lme_fit <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + prenyha + redo + dm + lv,
  random = ~ time | num,
  data = heart_jm2,
  na.action = na.exclude
)

cox_fit <- coxph(
  Surv(fuyrs, status) ~ hs + sex + age + bsa + lvh + prenyha + redo + dm + lv + cluster(num),
  data  = jm_data2,
  x = TRUE,
  model = TRUE
)

# JM using slope parameterization
joint_fit_slope <- jointModel(
  lmeObject = lme_fit,
  survObject = cox_fit,
  timeVar = "time",
  parameterization = "slope",
  method = "piecewise-PH-aGH",
  derivForm = list(
    fixed = ~ 1,
    indFixed = 2,   # "time" in fixef(lme_fit)
    random = ~ 1,
    indRandom = 2    # "time" random slope
  )
)
summary(joint_fit_slope)
```

What was done:

Data were cleaned to satisfy JM assumptions, including removal of duplicate
time points within subjects and exclusion of measurements occurring after the
event or censoring time. Because JM requires the same subjects in both the
longitudinal and survival components, we restricted the analysis to individuals
with complete covariate information. These steps ensured consistent risk-set
construction and stable estimation of the slope parameterization.

Moreover, the slope parameterization requires alignment of the derivative of
the longitudinal trajectory with the survival risk process. To avoid
zero-length intervals and dimension mismatch between the submodels, we enforced
strictly increasing measurement times within subjects and ensured identical
subject sets across the LME and Cox components. This alignment is necessary for
valid joint likelihood construction.

Interpretation of results:

In the JM with slope parameterization, the longitudinal submodel indicated that
LVMI declined modestly but significantly over time following surgery
($\beta$_time = -0.010, p = 0.044), with substantial between-patient
variability in baseline levels and slopes. In the survival submodel, older age
(p < 0.0001) and poorer preoperative ejection fraction (p = 0.0007) were
significantly associated with increased mortality risk. However, the
association parameter linking the instantaneous rate of change in LVMI to
mortality was not statistically significant ($\alpha$ = -6.23, p = 0.376),
suggesting that the rate of LVMI change does not independently predict
time-to-death in this cohort.

# comparison and conclusion

In the JM with value parameterization, the association parameter linking the
current level of LVMI to mortality was statistically significant ($\alpha$ =
1.28, p = 0.028). This indicates that higher LVMI levels are associated with
increased risk of death. Specifically, a one-unit increase in log(LVMI)
corresponds to an approximately 3.6-fold increase in hazard.

In contrast, the slope parameterization did not demonstrate a significant
association between the rate of LVMI change and mortality (p = 0.376). Model
comparison favored the value parameterization (AIC = 632.25 vs. 634.62),
suggesting that current LVMI level, rather than its rate of change, is more
strongly associated with time-to-death following aortic valve replacement.