---
title: "STAT 293 Final Project Modeling"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r,message=FALSE,warning=FALSE}
library(ggplot2)
library(survival)
library(nlme)
library(dplyr)
library(MASS)
library(JM)
```

```{r}
heart_data <- read.csv(file.choose())
```

```{r}
# clean and label the dataset to prep for modeling
heart_data$hs <- relevel(factor(heart_data$hs),
                         ref = "Stentless valve")

heart_data <- heart_data %>%
  mutate(num = factor(num),
         sex = factor(sex,
                      levels = c(0,1),
                      labels = c("Male", "Female")),
         prenyha = factor(prenyha),
         lvh = factor(lvh),
         redo = factor(redo),
         dm = factor(dm),
         lv = factor(lv))
```

# LMM full model

```{r}
lmm1 <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + prenyha + redo + dm + lv,
  random = ~ time | num,
  data = heart_data
  )
summary(lmm1)
```

## Residuals vs. fitted to check linearity and homoscedasticity

```{r}
lmm_df1 <- data.frame(fitted = fitted(lmm1),
                      resid = resid(lmm1,
                                    type = "pearson"))

ggplot(lmm_df1,
       aes(x = fitted,
           y = resid)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  labs(title = "LMM: Pearson Residuals vs. Fitted",
       x = "Fitted values",
       y = "Pearson residuals")
```

## Normal Q-Q plot of residuals to check approximate normality

```{r}
ggplot(lmm_df1,
       aes(sample = resid)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Pearson Residuals")
```

## Normal Q-Q plot of residuals of random effects to check normality and correlation

```{r}
re1 <- ranef(lmm1)

re_df1 <- data.frame(b0 = re1[, "(Intercept)"],
                     b1 = re1[, "time"])

ggplot(re_df1,
       aes(sample = b0)) +
  stat_qq(alpha = 0.6) + 
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Random Intercepts")

ggplot(re_df1,
       aes(sample = b1)) +
  stat_qq(alpha = 0.6) + 
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Random Slopes (time)")

ggplot(re_df1, 
       aes(x = b0,
           y = b1)) +
  geom_point(alpha = 0.6) +
  labs(title = "LMM: Random Effects (Intercept vs. Slope)",
       x = "Random intercept",
       y = "Random slope (time)")
```

## Check if nonlinear time is needed 

```{r}
lmm_quad <- update(lmm1,
                   fixed = . ~ . + I(time^2),
                   method = "REML")

anova(lmm1, lmm_quad)
```

# Cox PH full model

```{r}
# one row per patient
jm_data <- heart_data[!duplicated(heart_data$num), ]

cox1 <- coxph(
  Surv(fuyrs, status) ~ hs + sex + age + bsa + lvh + prenyha + redo + dm + lv +
    cluster(num),
  data = jm_data,
  x = TRUE,
  model = TRUE
)
summary(cox1)
```

## Schoenfeld residual test 

```{r}
ph1 <- cox.zph(cox1)

print(ph1)
plot(ph1)
```

## Functional form check for continuous predictors age

```{r, warning=FALSE, message=FALSE}
mart1 <- residuals(cox1,
                   type = "martingale")

cox_df1 <- data.frame(age = jm_data$age,
                      mart = mart1)

ggplot(cox_df1,
       aes(x = age,
           y = mart1)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(title = "Martingale Residuals vs. Age",
       x = "Age",
       y= "Martingale residuals")
```

# Joint full model

```{r}
joint1 <- jointModel(
  lmm1,
  cox1,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)
summary(joint1)
```

# Keep significant LMM covaraites

```{r}
lmm2 <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + dm + lv,
  random = ~ time | num,
  data = heart_data)
summary(lmm2)
```

## Residuals vs. fitted to check linearity and homoscedasticity

```{r}
lmm_df2 <- data.frame(fitted = fitted(lmm2),
                      resid = resid(lmm2,
                                    type = "pearson"))

ggplot(lmm_df2,
       aes(x = fitted,
           y = resid)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  labs(title = "LMM: Pearson Residuals vs. Fitted",
       x = "Fitted values",
       y = "Pearson residuals")
```

## Normal Q-Q plot of residuals to check approximate normality

```{r}
ggplot(lmm_df2,
       aes(sample = resid)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Pearson Residuals")
```

## Normal Q-Q plot of random effects to check normality and correlation

```{r}
re2 <- ranef(lmm2)

re_df2 <- data.frame(b0 = re2[, "(Intercept)"],
                     b1 = re2[, "time"])

ggplot(re_df2,
       aes(sample = b0)) +
  stat_qq(alpha = 0.6) + 
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Random Intercepts")

ggplot(re_df2,
       aes(sample = b1)) +
  stat_qq(alpha = 0.6) + 
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Random Slopes (time)")

ggplot(re_df2, 
       aes(x = b0,
           y = b1)) +
  geom_point(alpha = 0.6) +
  labs(title = "LMM: Random Effects (Intercept vs. Slope)",
       x = "Random intercept", y = "Random slope (time)")
```

## Check if nonlinear time is needed

```{r}
lme_quad2 <- update(lmm2,
                    fixed = . ~ . + I(time^2),
                    method = "REML")

anova(lmm2, lme_quad2)
```

# Perform stepwise on Cox PH

```{r}
cox_step <- stepAIC(cox1, direction = "backward")
summary(cox_step)
```

```{r}
cox2 <- coxph(Surv(fuyrs, status) ~ age + lv + cluster(num),
  data = jm_data,
  x = TRUE
)
summary(cox2)
```

## Schoenfeld residual test

```{r}
ph2 <- cox.zph(cox2)

print(ph2)
plot(ph2)
```

## Functional form check for continuous predictors age

```{r, warning=FALSE, message=FALSE}
mart2 <- residuals(cox2,
                   type = "martingale")

cox_df2 <- data.frame(age = jm_data$age,
                      mart2 = mart2)

ggplot(cox_df2,
       aes(x = age,
           y = mart2)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(title = "Martingale Residuals vs. Age",
       x = "Age",
       y= "Martingale residuals")
```

# Joint model 2

```{r}
joint2 <- jointModel(
  lmm2,
  cox2,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)
summary(joint2)

AIC(joint1, joint2)
```

# Keep significant Cox PH covaraites

```{r}
# drop lv2 
cox3 <- coxph(
  Surv(fuyrs, status) ~ age + I(lv == 3) + cluster(num),
  data = jm_data,
  x = TRUE,
  model = TRUE
)
summary(cox3)
```

## Schoenfeld residual test

```{r}
ph3 <- cox.zph(cox3)

print(ph3)
plot(ph3)
```

## Functional form check for continuous predictors age

```{r, warning=FALSE, message=FALSE}
mart3 <- residuals(cox3,
                   type = "martingale")

cox_df3 <- data.frame(age = jm_data$age,
                      mart3 = mart3)

ggplot(cox_df3,
       aes(x = age,
           y = mart3)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(title = "Martingale Residuals vs. Age",
       x = "Age",
       y= "Martingale residuals")
```

# Joint model 3

```{r}
joint3 <- jointModel(
  lmm2,
  cox3,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)
summary(joint3)

AIC(joint2, joint3)
```

# Prepare for joint model with slope/rate of change parameterization

```{r}
#reload the full dataset
heart_data <- read.csv(file.choose())
```

```{r}
# sort long data 
heart_sorted <- heart_data %>%
  arrange(num, time)

# build subject-level survival data from earliest row per subject
jm_data <- heart_sorted %>%
  distinct(num, .keep_all = TRUE) %>% # earliest time after sorting
  mutate(status = as.integer(status)) %>%
  filter(!is.na(fuyrs),
         fuyrs > 0,
         !is.na(status))

# longitudinal data for JM: drop subject-level survival columns to avoid duplicates
heart_long <- heart_sorted %>%
  dplyr::select(-any_of(c("fuyrs", "status"))) %>% # remove if exist in long file
  distinct(num, time, .keep_all = TRUE) # remove duplicate times

# enforce measurement times within follow-up 
heart_jm <- heart_long %>%
  inner_join(jm_data %>% dplyr::select(num, fuyrs),
             by = "num") %>%
  filter(time <= fuyrs) %>% # strict to avoid stop = start
  dplyr::select(-fuyrs)

# restrict to obs that complete data for both models
lmm_vars <- c("num", "time", "log.lvmi", "hs", "sex", "age", "bsa", "lvh", 
              "dm", "lv")
ids_lmm <- heart_jm %>%
  dplyr::select(all_of(lmm_vars)) %>%
  tidyr::drop_na() %>%
  distinct(num) %>%
  pull(num)

cox_vars <- c("num", "fuyrs", "status", "age", "lv")
ids_cox <- jm_data %>%
  dplyr::select(all_of(cox_vars)) %>%
  tidyr::drop_na() %>%
  distinct(num) %>%
  pull(num)

keep_ids <- intersect(ids_lmm, ids_cox)
heart_jm2 <- heart_jm %>% filter(num %in% keep_ids)
jm_data2 <- jm_data %>% filter(num %in% keep_ids)
```

## LMM using the cleaned dataset

```{r}
lmm3 <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + dm + lv,
  random = ~ time | num,
  data = heart_jm2
)
summary(lmm3)
```

### Residuals vs. fitted to check linearity and homoscedasticity

```{r}
lmm_df3 <- data.frame(fitted = fitted(lmm3),
                      resid = resid(lmm3,
                                    type = "pearson"))

ggplot(lmm_df3,
       aes(x = fitted,
           y = resid)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  labs(title = "LMM: Pearson Residuals vs. Fitted",
       x = "Fitted values",
       y = "Pearson residuals")
```

### Normal Q-Q plot of residuals to check approximate normality

```{r}
ggplot(lmm_df3,
       aes(sample = resid)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Pearson Residuals")
```

### Normal Q-Q plot of random effects to check normality and correlation

```{r}
re3 <- ranef(lmm3)

re_df3 <- data.frame(b0 = re3[, "(Intercept)"],
                     b1 = re3[, "time"])

ggplot(re_df3,
       aes(sample = b0)) +
  stat_qq(alpha = 0.6) + 
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Random Intercepts")

ggplot(re_df3,
       aes(sample = b1)) +
  stat_qq(alpha = 0.6) + 
  stat_qq_line() +
  labs(title = "LMM: Q-Q Plot of Random Slopes (time)")

ggplot(re_df3, 
       aes(x = b0,
           y = b1)) +
  geom_point(alpha = 0.6) +
  labs(title = "LMM: Random Effects (Intercept vs Slope)",
       x = "Random intercept", y = "Random slope (time)")
```

### Check if nonlinear time is needed 

```{r}
lme_quad3 <- update(lmm3,
                    fixed = . ~ . + I(time^2),
                    method = "REML")

anova(lmm3, lme_quad3)
```

## Cox PH using the cleaned dataset

```{r}
cox4 <- coxph(
  Surv(fuyrs, status) ~ age + I(lv == 3) + cluster(num),
  data = jm_data2,
  x = TRUE
)
summary(cox4)
```

### Schoenfeld residual test

```{r}
ph4 <- cox.zph(cox4)

print(ph4)
plot(ph4)
```

### Functional form check for continuous predictors age

```{r, warning=FALSE, message=FALSE}
mart4 <- residuals(cox4,
                   type = "martingale")

cox_df4 <- data.frame(age = jm_data$age,
                      mart4 = mart4)

ggplot(cox_df4,
       aes(x = age,
           y = mart4)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(title = "Martingale Residuals vs. Age",
       x = "Age",
       y= "Martingale residuals")
```

# Joint model with slope/rate of change parameterization

```{r}
joint_slope <- jointModel(
  lmeObject = lmm3,
  survObject = cox4,
  timeVar = "time",
  parameterization = "slope",
  method = "piecewise-PH-aGH",
  derivForm = list(
    fixed = ~ 1, # derivative of fixed part wrt time = intercept column
    indFixed = 2, # index of time in fixef(lme_fit) 
    random = ~ 1, # derivative of random part wrt time
    indRandom = 2 # index of time in the random effects
  )
)
summary(joint_slope)

AIC(joint3, joint_slope)
```

# Sensitivity analysis

## Extending Cox PH by adding more baseline covariates

```{r}
cox_sens <- coxph(
  Surv(fuyrs, status) ~ age + I(lv == 3) + sex + hs + cluster(num),
  data = jm_data2,
  x = TRUE,
  model = TRUE
)
summary(cox_sens)
```

### Schoenfeld residual test

```{r}
ph_sens <- cox.zph(cox_sens)

print(ph_sens)
plot(ph_sens)
```

### Functional form check for continuous predictors age

```{r, warning=FALSE, message=FALSE}
mart_sens <- residuals(cox_sens,
                       type = "martingale")

cox_df_sens <- data.frame(age = jm_data$age,
                          mart_sense = mart_sens)

ggplot(cox_df_sens,
       aes(x = age,
           y = mart_sens)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(title = "Martingale Residuals vs. Age",
       x = "Age",
       y= "Martingale residuals")
```

## Joint model sensitivity analysis 1

```{r}
joint_sens1 <- jointModel(
  lmeObject = lmm3,
  survObject = cox_sens,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)
summary(joint_sens1)   

AIC(joint3, joint_sens1)
```

## Joint model sensitivity analysis 2

```{r}
# use value + slope for parameterization to test whether adding slope on top of
# value improves fit
joint_sens2 <- jointModel(
  lmeObject = lmm3,
  survObject = cox4,
  timeVar = "time",
  method = "piecewise-PH-aGH",
  parameterization = "both",
  derivForm = list(
    fixed = ~ 1, # derivative of fixed part wrt time = intercept column
    indFixed = 2, # index of time in fixef(lme_fit) 
    random = ~ 1, # derivative of random part wrt time
    indRandom = 2 # index of time in the random effects
  )
)
summary(joint_sens2)  

AIC(joint3, joint_sens2) 
```