---
title: "STAT 293 - Final Project"
author: "Nancy Lopez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(ggplot2)
library(mice)
library(dplyr)
library(naniar)
library(survival)
library(lme4)
```


```{r}
load("/Users/nancylopez/Downloads/heart.rdata")
```

# MISSING DATA

```{r}
missing_table <- data.frame(
  variables = names(heart_data),
  n_missing = colSums(is.na(heart_data)),
  percentage_missing = colMeans(is.na(heart_data)) * 100
)

missing_table[order(-missing_table$percentage_missing), ]

```


```{r}
missing_data <- heart_data[c("log.grad", "grad", "ef", "lvmi")]
md.pattern(missing_data)

```


```{r}
miss_data <- heart_data
miss_data$lvmi_miss <- is.na(heart_data$lvmi)
with(miss_data, table(status, lvmi_miss))
prop.table(with(miss_data, table(status, lvmi_miss)), 1)
```

For alive patients, 16.1 percent of the lmvi observations are missing
For dead patients, 20.5 percent of the lmvi observations are dead

```{r}
miss_data$ef_miss <- is.na(heart_data$ef)
with(miss_data, table(status, ef_miss))
prop.table(with(miss_data, table(status, ef_miss)),1)
```

not much of a diff between dead patients and lost at follow up, so we lean more towards MAR than MNAR 

```{r}
miss_data$grad_miss <- is.na(heart_data$grad)
with(miss_data, table(status, grad_miss))
prop.table(with(miss_data, table(status, grad_miss)), 1)
```

most missing observations are alive, since we have like a 8% advantage from the live portion, we are more likely to assume MAR

## MCAR Test 

```{r}
mcar_test(heart_data)

# p > 0.05 == MCAR
# P < 0.05 == not MCAR 

# actual p: 3.595806e-05	
```

so now we know fs its prob not MCAR

"Little's missing completely at random (MCAR) test" 

## Visits 

```{r}
# number of patients
n_patients <- length(unique(heart_data$num))

# total observations
n_obs <- nrow(heart_data)

# visits per patient
visits_summary <- heart_data %>%
  group_by(num) %>%
  summarise(n_visits = n()) %>%
  summary()

visits_per_patient <- heart_data %>%
  group_by(num) %>%
  summarise(n_visits = n())

n_patients
n_obs
visits_summary
```

```{r}
visits_per_patient
table(visits_per_patient$n_visits)
```

```{r}

data_mi <- heart_data


# run multiple imputation
imp <- mice(
  data_mi,
  m = 20,
  seed = 293
)

summary(imp)
```


# EDA

## Baselline table

```{r}
# pick baseline vars (edit if you want)
key_vars <- heart_data %>%
  select(sex, age, time, fuyrs, status, grad, log.lvmi, ef, lvh, prenyha, redo, size, con.cabg, creat, dm, acei, lv, emergenc, hc, sten.reg.mix, hs)

baseline_summary <- key_vars %>%
  group_by(hs) %>%
  summarise(across(everything(),
                   list(mean = ~mean(. , na.rm = TRUE),
                        sd = ~sd(. , na.rm = TRUE))))

baseline_summary
```

## Boxplots of selected baselines covariates by survival status 

```{r}
ggplot(heart_data, aes(x = factor(status), y = sex)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Sex")

ggplot(heart_data, aes(x = factor(status), y = age)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Age")

ggplot(heart_data, aes(x = factor(status), y = grad)) +
  geom_boxplot() + 
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Valve gradient at follow-up")

ggplot(subset(heart_data, grad < 30), 
       aes(x = factor(status), y = grad)) +
  geom_boxplot() + 
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Valve gradient at follow-up")

ggplot(heart_data, aes(x = factor(status), y = lvmi)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Left ventricular mass index")

ggplot(subset(heart_data, lvmi < 300), aes(x = factor(status), y = lvmi)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Left ventricular mass index")

ggplot(heart_data, aes(x = factor(status), y = ef)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Ejection fraction at follow-up visit")

ggplot(heart_data, aes(x = factor(status), y = bsa)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative body surface area")

ggplot(heart_data, aes(x = factor(status), y = size)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "size of the valve (millimeters)")

ggplot(heart_data, aes(x = factor(status), y =  creat)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative serum creatinine")

ggplot(subset(heart_data, creat < 200), aes(x = factor(status), y =  creat)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative serum creatinine")

ggplot(heart_data, aes(x = factor(status), y =  creat)) +
  geom_boxplot() +
  labs(x = "Survival status (0=alive,1=dead)",
       y = "Preoperative serum creatinine")
```
- age 
- lmvi
- bsa

## Missingness patterns vs. survival status plot

```{r}
heart_data$lvmi_miss <- is.na(heart_data$lvmi)

ggplot(heart_data, aes(x = factor(status), fill = factor(lvmi_miss))) +
  geom_bar(position = "fill") +
  labs(x ="Survival status (0=alive,1=dead)", fill = "Is data missing?")
```

## Spaghetti plot of log.lvmi over time

```{r}
ggplot(heart_data, aes(x = time, y = log.lvmi, group = num)) +
  geom_line(alpha = 0.2) +
  labs(x = "Time since surgery (years)",
       y = "log(LVMI)")
```
Individuals in log(LVMI) show substantial heterogeneity, supporting the use of a mixed-effects model for the longitudinal component???

## Kaplan-Meier curves by valve type

```{r}
# create a survival object
# fuyrs = follow-up time
surv_obj <- Surv(heart_data$fuyrs, heart_data$status)

# this estimates a separate survival curve for each valve group
fit <- survfit(surv_obj ~ hs, data = heart_data)

plot(fit,
     col = c("blue", "red"),
     xlab = "Time since surgery (years)",
     ylab = "Survival probability")

legend("bottomleft",
       legend = c("Porcine", "Homograft"),
       col = c("blue", "red"),
       lty = 1)
```
Potential differences in survival between valve types, with the homograft group showing lower survival probabilities over follow-up. This motivates inclusion of valve type in the survival submodel

```{r,warning=FALSE,message=FALSE}
ggplot(heart_data,
       aes(time, log.lvmi, color = factor(status))) +
  geom_smooth(se=FALSE)
```

# ANALYSIS DRAFT

```{r}
lmm_fit <- lmer(
  log.lvmi ~ time * hs + sex + age + bsa + lvh + prenyha + redo + dm + lv + (time | num),
  data = heart_data)

summary(lmm_fit)
```

WHAT I GOT FOR LMM:
- interaction: 0.005 No evidence that rate of LVMI change differs by valve type
- time: -0.012 LVMI slighlty decreases over time after surgery 
- hs: 0.068 Patients with stentless valves have slightly higher LVMI, but not statistically convincing.
- sex: -0.218 Females have lower LVMI on average
- age: 0.001 not significant
- bsa: -0.274 Larger BSA is associated with lower indexed LVMI
- lvh: -0.114 patients with preoperative LVH have slightly lower postoperative LVMI
- prenyha: -0.005 not significant
- redo: -0.007 Patients with prior surgery have slightly higher LVMI, but effect is small
- dm: -0.191 slightly lower LVMI among diabetes 
- lv: 0.038 higher lvef means higher lmvi

```{r}
survival_fit <- coxph(
  Surv(fuyrs, status) ~ hs +
    sex + age + bsa + lvh + prenyha +
    redo + dm + lv,
  data = heart_data
)

summary(survival_fit)
```
WHAT I GOT FOR SURVIVAL ANALYSIS:
- hr: 2.02 patients with stentless valves have about 2× hazard of death
- age: 1.18 each additional year of age means 18% higher hazard 
- lvh: 0.14 
- redo 0.494 prior surgery associated with lower observed hazard
- lv: 2.49 worse LV function means higher mortality risk


# ANALYSIS BASED OF ARTICLE WITH JM

```{r}
#install.packages("nlme")
library(nlme)
#install.packages("JM")
library(JM)

jm_data <- heart_data[!duplicated(heart_data$num), ]

lme_fit <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + prenyha + redo + dm + lv,
  random = ~ time | num,
  data = heart_data)

cox_fit <- coxph(
  Surv(fuyrs, status) ~ hs + sex + age + bsa + lvh + prenyha + redo + dm + lv +
    cluster(num),
  data = jm_data,
  x = TRUE,
  model = TRUE
)


joint_fit <- jointModel(
  lme_fit,
  cox_fit,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)

summary(joint_fit)
```

