---
title: "STAT293_FinalProject_Model"
author: "Ryan Kitagawa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# load libraries
library(mice)
library(dplyr)
library(nlme)
library(lme4)
library(lmerTest)
library(survival)
library(JMbayes2)
```


```{r}
# load data
load("/Users/rjkit/Downloads/heart.rdata")
```


Do model here

```{r}
# convert categorical variables to factors
heart_data <- heart_data %>%
  mutate(
    sex = as.factor(sex),
    prenyha = as.factor(prenyha),
    redo = as.factor(redo),
    con.cabg = as.factor(con.cabg),
    dm = as.factor(dm),
    acei = as.factor(acei),
    lv = as.factor(lv),
    emergenc = as.factor(emergenc),
    hc = as.factor(hc),
    sten.reg.mix = as.factor(sten.reg.mix),
    hs = as.factor(hs) # Your primary predictor
  )
```


Longitudinal submodel
```{r}
# base submodel here
long_sm <- lme(fixed = log.lvmi ~ time * hs + sex + age + bsa + lvh + prenyha +
  redo + size + con.cabg + creat + dm + acei + lv + emergenc + hc +
  sten.reg.mix, random = ~ time | num, data = heart_data)

# summary of model
summary(long_sm)

long_sm_lmer <- lmer(log.lvmi ~ time * hs + sex + age + bsa + lvh + prenyha + redo + 
  size + con.cabg + creat + dm + acei + lv + emergenc + hc + sten.reg.mix + 
  (time | num), data = heart_data)

# summary of  lmer model for sanity check

# stepwise to see initial changes
step(long_sm_lmer, reduce.random = FALSE, direction = "backward")

# save reduced model with what stepwise found
long_red_sm <- lme(fixed = log.lvmi ~ time + hs + sex + bsa + size + lv,
    random = ~ time | num, data = heart_data)

# summary of reduced model
summary(long_red_sm)
```


Survival Submodel

```{r}
# convert into wide format for survival
heart_data_surv <- heart_data[!duplicated(heart_data$num), ]

# base submodel here
surv_sm <- coxph(Surv(fuyrs, status) ~ hs + sex + age + bsa + lvh +
    prenyha + redo + size + con.cabg + creat + dm + acei + lv + 
    emergenc + hc + sten.reg.mix, data = heart_data_surv, x = TRUE, 
    model = TRUE)


# summary of model here
summary(surv_sm)


# initial stepwise here
step(surv_sm, direction = "backward")

# save reduced model
surv_red_sm <- coxph(formula = Surv(fuyrs, status) ~ age + lvh + 
  con.cabg + lv + hc + sten.reg.mix, data = heart_data_surv, 
  model = TRUE, x = TRUE)

# summarize model
summary(surv_red_sm)
```


Joint model

```{r}
# with JMbayes2 (not needed?)

# both full models for joint model
joint_full <-jm(surv_sm, long_sm, time_var = "time")

# summary
summary(joint_full)

# both reduced models for joint model
joint_red <- jm(surv_red_sm, long_red_sm, time_var = "time")

# summary
summary(joint_red)
```

```{r}
# using JM package
library(JM)

# full model
jm_full <- jointModel(
  long_sm,
  surv_sm,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)

# summary of full model
summary(jm_full)

# reduced model
jm_red <- jointModel(
  long_red_sm,
  surv_red_sm,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)

# summary of reduced mdoel
summary(jm_red)


```


# ANALYSIS DRAFT

```{r}
lmm_fit <- lmer(
  log.lvmi ~ time * hs + sex + age + bsa + lvh + prenyha + redo + dm + lv + (time | num),
  data = heart_data)

summary(lmm_fit)
```

WHAT I GOT FOR LMM:
- interaction: 0.005 No evidence that rate of LVMI change differs by valve type
- time: -0.012 LVMI slighlty decreases over time after surgery 
- hs: 0.068 Patients with stentless valves have slightly higher LVMI, but not statistically convincing.
- sex: -0.218 Females have lower LVMI on average
- age: 0.001 not significant
- bsa: -0.274 Larger BSA is associated with lower indexed LVMI
- lvh: -0.114 patients with preoperative LVH have slightly lower postoperative LVMI
- prenyha: -0.005 not significant
- redo: -0.007 Patients with prior surgery have slightly higher LVMI, but effect is small
- dm: -0.191 slightly lower LVMI among diabetes 
- lv: 0.038 higher lvef means higher lmvi

```{r}
survival_fit <- coxph(
  Surv(fuyrs, status) ~ hs +
    sex + age + bsa + lvh + prenyha +
    redo + dm + lv,
  data = heart_data
)

summary(survival_fit)
```
WHAT I GOT FOR SURVIVAL ANALYSIS:
- hr: 2.02 patients with stentless valves have about 2× hazard of death
- age: 1.18 each additional year of age means 18% higher hazard 
- lvh: 0.14 
- redo 0.494 prior surgery associated with lower observed hazard
- lv: 2.49 worse LV function means higher mortality risk


# ANALYSIS BASED OF ARTICLE WITH JM

```{r}
#install.packages("nlme")
library(nlme)
#install.packages("JM")
library(JM)

jm_data <- heart_data[!duplicated(heart_data$num), ]

lme_fit <- lme(
  fixed = log.lvmi ~ time + hs + sex + age + bsa + lvh + prenyha + redo + dm + lv,
  random = ~ time | num,
  data = heart_data)

cox_fit <- coxph(
  Surv(fuyrs, status) ~ hs + sex + age + bsa + lvh + prenyha + redo + dm + lv +
    cluster(num),
  data = jm_data,
  x = TRUE,
  model = TRUE
)


joint_fit <- jointModel(
  lme_fit,
  cox_fit,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)

summary(joint_fit)
```

